Self-assembly of colloid-cholesteric composites: A route to switchable optical materials

### 2013-10-02 trunk/papers/2012-sandwich/support.tex
Updated supporting text and figures
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2164 40876441-3496-0410-a613-a2fca35b2ae2

In this work, the chirality and reduced temperature are always set to lie
within the blue phase I region of the equilibrium phase diagram, except
for Fig. 3 in the main manuscript, where the switching of colloids in
blue phase III is studied.

In addition to the fluid free density density $f(Q_{\alpha\beta})$, a surface
free energy density is also present to represent the energetic cost of
anchoring at a solid surface. If the preferred order parameter tensor
at the solid surface is $Q^0_{\alpha\beta}$, then the surface free energy
density (per unit area) is
\begin{equation}
f_s(Q_{\alpha\beta}, Q^0_{\alpha\beta})
= {\textstyle \frac{1}{2}}W(Q_{\alpha\beta} - Q^0_{\alpha\beta})^2,
\end{equation}
where $W$ is a constant determining the strength of the anchoring.
The determination of $Q^0_{\alpha\beta}$ for normal and planar anchoring is
discussed below. For a colloid of radius $a$, the strength of the surface
anchoring compared which the bulk fluid elastic constant may be quantified
by the dimensionless parameter $Wa/K$. For small values of this parameter,
the presence of a colloid surface should have little impact on the local
fluid LC ordering.

The effect of an electric field can be modelled by including the following
term in the bulk free energy density:
\begin{equation}
-\frac{\epsilon_a}{12\pi} E_{\alpha}Q_{\alpha\beta}E_{\beta},
\end{equation} 
where $E_{\alpha}$ are the components of the electric field, and 
$\epsilon_a$ (here assumed positive) is the dielectric anisotropy of the liquid
crystal. The strength of the electric field is quantified via
a further dimensionless number, 
\begin{equation}
{\cal E}^2 = \frac{27 \epsilon_a}{32 \pi A_0 \gamma} E_\alpha E_\alpha.
\end{equation}

@@

The momentum evolution obeys a Navier-Stokes equation driven by the divergence
of a generalised stress $P_{\alpha\beta}$:
\begin{equation}
\label{nse}
\rho\,\partial_tu_\alpha +\rho \,u_\beta \partial_\beta u_\alpha
=\partial_\beta P_{\alpha\beta}
\end{equation}
The pressure tensor $P_{\alpha\beta}$ is, in general, asymmetric and includes
both viscous and thermodynamic components:
\begin{eqnarray}
P_{\alpha \beta}&=&
- p_0 \delta_{\alpha \beta} 
+ \eta \{ \partial_\alpha u_\beta + \partial_\beta u_\alpha\}
\nonumber\\
&-&  \xi H_{\alpha \gamma}\left(Q_{\gamma \beta}
+ {\textstyle \frac{1}{3}} \delta_{\gamma \beta} \right)
-\xi \left(Q_{\alpha \gamma}
+ { \textstyle \frac{1}{3}} \delta_{\alpha \gamma}\right) H_{\gamma \beta}
\nonumber\\ 
&+& 2 \xi  \left(Q_{\alpha \beta}
+ {\textstyle \frac{1}{3}} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}
-\partial_\alpha Q_{\gamma \nu}
\frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
&+& Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta}.
\end{eqnarray}
A lattice Boltzmann flow solver is used, where the isotropic pressure $p_0$
and viscous terms are managed directly by the solver (as in a simple Newtonian
fluid, of viscosity $\eta$), whereas the divergence of the remaining terms
is treated as a local force density on the fluid. In all simulations
reported the mean fluid density $\rho = 1$ and the viscosity
$\eta = 0.01$ (simulations in confined geometry also used enhanced
bulk viscosity $\zeta = 10\eta$ to ensure numerical stability).

@@

Hydrodynamic boundary conditions for solid objects are handled via
the lattice Boltzmann solver. In particular, the standard method of
bounce-back on links is used for both colloids and confining walls
\cite{ladd94,nguyen2002}. Boundary conditions for the order parameter
are set out in the following.

@@

The following protocol was used for the simulations of BPIII with
colloidal particles: First, we started from a configuration of
randomly oriented and positioned double twist cylinders from which an
amorphous BPIII network emerged. The structure
was equilibrated for $6\times10^{5}$ LB time steps until the relaxation came
virtually to a standstill. Then the colloidal particles featuring week
normal anchoring were randomly inserted and the composite system was
equilibrated for another $1.2\times10^{6}$ time steps.
At the end of the equilibration phase the homogenous electric field
with reduced field strength $e=\sqrt{\frac{27\epsilon_a}{32\pi A_0\gamma}} E=0.8$ 
was switched on along the z-direction for $1.2\times10^{6}$ time steps.
This was found to be sufficiently long for the hexagonal structure to emerge and saturate.
After the electric field has been switched off the system quickly relaxed
to a metastable state which showed a residual anisotropy in the
colloid distribution. The total length of this final relaxation phase
was $1\times10^6$ LB time steps.

@@

To get from simulation to physical units, we need to calibrate scales of
length, energy and time in the simulations. For similar mappings see,
e.g.~\cite{denniston2}.

The length scale can be set by mapping our half pitch to a realistic blue
phase unit cell size, which are in the 100-500 nm range \cite{blue1}. 
For bulk simulations (Fig.~1 and Fig.~3 in the main text), a suitable choice
is one where one simulation unit (lattice site) corresponds to 10~nm. Therefore
the colloidal size in Figs.~1 and~3 corresponds to $\sim$ 50~nm.

To obtain an energy scale, we choose $A_0 \simeq 10^6$~Pa, which is 
reasonable from Ref.~\cite{blue1} (this is also the choice in
Ref.~\cite{oliver2}). This choice leads to a simulation unit of
free energy (or stress) equal to $10^{8}$~Pa in Figs.~1 and~3.
From the energy and length scales one can map elastic constants to
physical units: for instance, the simulation in Fig.~1 corresponds to
a liquid crystal with splay, bend and twist (Frank) elastic constants
equal to 35 pN (or equivalently $K=$70~pN). That is,
WHAT ARE a/b???

For the timescale calibration we use the formula (see e.g.~\cite{denniston})
\begin{equation}
\gamma_1=\frac{2q^2}{\Gamma}
\end{equation}
which relates the rotational viscosity $\gamma_1$, to the ordering strength
$q$ and the order parameter mobility $\Gamma$.
In our simulations, we choose $\Gamma = 0.5$, while the value of $\gamma$
used in Fig.~1 leads to $q=1/2$.   
For real liquid crystalline materials, $\gamma_1$ usually lies in range
$10^{-2}-1$ Pa s~\cite{deGennes}; we choose for definiteness
$\gamma_1 = 1$~Pa~s (equivalently 10 poise). Given the previous mapping for
free energy (or stress) units, this leads to a simulation unit of time equal
to $10^{-8}$~s. The total simulated time in Figs. 1-3 is therefore in the
millisecond range.

To calibrate the electric field strength 
we equate the value of the dimensionless parameter ${\cal E}$ in our 
simulations with that of a hypothetical experiment.
Given the previous value for $A_0$, and a dielectric anisotropy of 20
(or equivalently $\epsilon_a=20\epsilon_0$, with $\epsilon_0$ the
dielectric permittivity of vacuum), we then obtain that an electric field
of 100 V/$\mu$m corresponds to a value of ${\cal E}\sim 0.4$ (considering
a value of $\gamma=3.0857$ as in Fig. 3 of the main text.

Given these mappings, all quantities can be easily transformed from
simulation to physical units and vice versa. For example Figs.~1 and~2
use an isotropic viscosity $\eta \simeq$~0.01 and~0.1 in simulation units,
which translate into $0.01-0.1$~Pa~s respectively. This is sensible (if 
slightly low) for a molecular nematogen in the isotropic phase. 
The effective viscosity in ordered phases is higher, but this is ensured
in our model by the coupling to the order parameter.
We note that the density is set to unity in our simulations: this
corresponds to a fluid density which is about a thousand times larger
than in experiments. This however makes no difference in practice as
the Reynolds number $= \rho V\Lambda/\eta$ (where $V$ is a typical
velocity, in our simulations around $10^{-5}$ in simulation units) 
remains small enough~\cite{codef}.

To summarize, our simulations represent BP-forming materials, with the
interpretation of the simulation units for length, time, and energy
density being close to 10~nm, 10~ns, and 100~MPa respectively.

@@

%ISSUE: do we want some free energy numbers?

@@

@@

@@

%\vfill\pagebreak

@@

\if{\begin{figure}[!h]
\begin{center}
%%\includegraphics[width=0.95\columnwidth]{s5.jpg}
\end{center}
\caption{Pair correlation function for weak ($WR/K=0.23$) and strong ($WR/K=2.3$) anchoring cases considered in the manuscript. (A,B) BPI initialisation with A) volume fraction $\phi=1${\%} and B) $\phi=5${\%}. (C,D) ``quenched system'' with C) $\phi=1${\%} and D) $\phi=5${\%}.}
\end{figure}
}\fi

@@

{\bf Supplementary Movie 1:} 
Final state of the end of the simulation in Fig.~1A. The movie
shows a rotation of the configuration to better appreciate details.\\

{\bf Supplementary Movie 2:} 
Final state of the end of the simulation in Fig.~1B. The movie
shows a rotation of the configuration to better appreciate details.\\

{\bf Supplementary Movie 3:} 
Final state of the end of the simulation in Fig.~1C. The movie
shows a rotation of the configuration to better appreciate details. \\

{\bf Supplementary Movie 4:} 
Final state of the end of the simulation in Fig.~1D. The movie
shows a rotation of the configuration to better appreciate details. \\

{\bf Supplementary Movie 5:} 
Final state at the end of the initial relaxation phase in Fig.~3. The movie
shows a rotation of the configuration to better appreciate details. The distribution of the colloidal particles is isotropic with the particles forming clusters and having a tendency to occupy positions a the intersection of the BPIII disclination lines.\\

{\bf Supplementary Movie 6:} 
Configuration in a homogeneous electric field with strength
corresponding to ${\cal E}=0.8$. The movie
shows a rotation of the configuration to better appreciate details.
A hexagonal pattern emerges where the particles form chains
situated at the vertices of the honeycomb-like structure. \\

>>>>

The effect of an electric field can be modelled by including the following
term in the bulk free energy density:
\begin{equation}
-\frac{\epsilon_a}{12\pi} E_{\alpha}Q_{\alpha\beta}E_{\beta},
\end{equation} 
where $E_{\alpha}$ are the components of the electric field, and 
$\epsilon_a$ (here assumed positive) is the dielectric anisotropy of the liquid
crystal. The strength of the electric field is quantified via
one further dimensionless number
\begin{equation}
{\cal E}^2 = \frac{27 \epsilon_a}{32 \pi A_0 \gamma} E_\alpha E_\alpha.
\end{equation}
The quantity ${\cal E}$ will be referred to as the reduced field strength.

{\bf 1.1 Surface free energy}

In addition to the fluid free energy density $f(Q_{\alpha\beta})$, a surface
free energy density is also present to represent the energetic cost of
anchoring at a solid surface. If the preferred order parameter tensor
at the solid surface is $Q^0_{\alpha\beta}$, then the surface free energy
density (per unit area) is
\begin{equation}
f_s(Q_{\alpha\beta}, Q^0_{\alpha\beta})
= {\textstyle \frac{1}{2}}W(Q_{\alpha\beta} - Q^0_{\alpha\beta})^2,
\end{equation}
where $W$ is a constant determining the strength of the anchoring.
The determination of $Q^0_{\alpha\beta}$ for normal and planar anchoring is
discussed below. For a colloid of radius $a$, the strength of the surface
anchoring compared which the bulk fluid elastic constant may be quantified
by the dimensionless parameter $Wa/K$. For small values of this parameter,
the presence of a colloid surface should have little impact on the local
fluid LC ordering.

@@

The momentum evolution obeys a Navier-Stokes equation driven by the divergence
of a generalised stress $P_{\alpha\beta}$:
\begin{equation}
\label{nse}
\rho\,\partial_tu_\alpha +\rho \,u_\beta \partial_\beta u_\alpha
=\partial_\beta P_{\alpha\beta}.
\end{equation}
The pressure tensor $P_{\alpha\beta}$ is, in general, asymmetric and includes
both viscous and thermodynamic components:
\begin{eqnarray}
P_{\alpha \beta}&=&
- p_0 \delta_{\alpha \beta} 
+ \eta \{ \partial_\alpha u_\beta + \partial_\beta u_\alpha\}
\nonumber\\
&-&  \xi H_{\alpha \gamma}\left(Q_{\gamma \beta}
+ {\textstyle \frac{1}{3}} \delta_{\gamma \beta} \right)
-\xi \left(Q_{\alpha \gamma}
+ { \textstyle \frac{1}{3}} \delta_{\alpha \gamma}\right) H_{\gamma \beta}
\nonumber\\ 
&+& 2 \xi  \left(Q_{\alpha \beta}
+ {\textstyle \frac{1}{3}} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}
-\partial_\alpha Q_{\gamma \nu}
\frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
&+& Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta}.
\end{eqnarray}
A lattice Boltzmann flow solver is used, where the isotropic pressure $p_0$
and viscous terms are managed directly by the solver (as in a simple Newtonian
fluid, of viscosity $\eta$), whereas the divergence of the remaining terms
is treated as a local force density on the fluid. In all simulations
reported the mean fluid density $\rho = 1$ and the viscosity
$\eta = 0.01$ (simulations in confined geometry also used enhanced
bulk viscosity $\zeta = 10\eta$ to ensure numerical stability).

@@

Hydrodynamic boundary conditions for solid objects are handled via
the lattice Boltzmann solver. In particular, the standard method of
bounce-back on links is used for both colloids and confining walls
\cite{ladd94,nguyen2002}. Boundary conditions for the order parameter
are set out in \cite{juho}.

@@

The following protocol is used for the simulations of BPIII with
colloidal particles. An initial configuration of
randomly oriented and positioned double twist cylinders is used,
from which an amorphous BPIII network emerges \cite{oliver-bp3}. This structure
is equilibrated for $6\times10^{5}$ LB time steps until no significant
further evolution is observed. The colloidal particles, with weak
normal anchoring, are then inserted with random chosen positions,
and the composite system is equilibrated for another $1.2\times10^{6}$
time steps.
At the end of the equilibration phase the uniform electric field
with reduced field strength ${\cal E} = 0.8$ 
is switched on (oriented along the $z$-direction), and the simulation
run for a further $1.2\times10^{6}$ time steps.
This is found to be sufficiently long for a hexagonal structure to
emerge and saturate.
If the electric field is then switched off, and the system quickly relaxes
to a metastable state which shows a residual anisotropy in the
colloidal distribution. The total length of this final relaxation phase
is $1\times10^6$ LB time steps.

@@

To get from simulation to physical units, a calibration of scales of
length, energy and time in the simulations is required. For similar
mappings see,
e.g.~\cite{denniston2}.

The length scale can be set by mapping the half pitch to a realistic blue
phase unit cell size, which is in the 100--500~nm range \cite{blue1}. 
For bulk simulations (Fig.~1 and Fig.~3 in the main text), a suitable choice
is one where one simulation unit (lattice site) corresponds to 10~nm. Therefore
the colloidal size in Figs.~1 and~3 corresponds to $\sim$ 50~nm.

To obtain an energy scale, it is possible to choose $A_0 \simeq 10^6$~Pa,
which is reasonable following Ref.~\cite{blue1} (this is also the choice in
Ref.~\cite{oliver2}). This choice leads to a simulation unit of
free energy (or stress) equal to $10^{8}$~Pa in Figs.~1 and~3.
From the energy and length scales one can map elastic constants to
physical units: for instance, the simulation in Fig.~1 corresponds to
a liquid crystal with splay, bend and twist (Frank) elastic constants
equal to 35 pN (or equivalently $K=70$~pN).

The timescale calibration may be obtained from (see e.g.~\cite{denniston})
\begin{equation}
\gamma_1=\frac{2q^2}{\Gamma},
\end{equation}
which relates the rotational viscosity $\gamma_1$ to the ordering strength
$q$ and the order parameter mobility $\Gamma$.
In the present simulations, $\Gamma = 0.5$ and the value of $\gamma$
for the simulation in Fig.~1 leads to $q=1/2$.   
For real liquid crystalline materials, $\gamma_1$ usually lies in range
$10^{-2}-1$ Pa s~\cite{deGennes}; for definiteness say
$\gamma_1 = 1$~Pa~s (equivalently 10 poise). Given the previous mapping for
free energy (or stress) units, this leads to a simulation unit of time equal
to $10^{-8}$~s. The total simulated time in Figs. 1-3 is therefore in the
millisecond range.

To calibrate the electric field strength 
one can equate the value of the dimensionless parameter ${\cal E}$ in the 
simulations with that of a hypothetical experiment.
Given the previous value for $A_0$, and a dielectric anisotropy of 20
(or equivalently $\epsilon_a=20\epsilon_0$, with $\epsilon_0$ the
dielectric permittivity of vacuum), an electric field
of 100 V/$\mu$m is obtained. This corresponds to a value of
${\cal E}\sim 0.4$ (considering a value of $\gamma=3.0857$ as in
Fig.~3 of the main text).

Given these mappings, all quantities can be easily transformed from
simulation to physical units and vice versa. For example Figs.~1 and~2
use an isotropic viscosity $\eta \simeq$~0.01 and~0.1 in simulation units,
which translate into $0.01-0.1$~Pa~s respectively. This is sensible (if 
slightly low) for a molecular nematogen in the isotropic phase. 
The effective viscosity in ordered phases is higher, but this is ensured
in our model by the coupling to the order parameter.
Note that the density is set to unity in the simulations: this
corresponds to a fluid density which is about a thousand times larger
than in experiments. This makes no difference in practice as
the Reynolds number $= \rho V\Lambda/\eta$ (where $V$ is a typical
velocity, in our simulations around $10^{-5}$ in simulation units) 
remains small enough~\cite{codef}.

To summarize, the simulations represent BP-forming materials, with the
interpretation of the simulation units for length, time, and energy
density being close to 10~nm, 10~ns, and 100~MPa respectively.

@@

@@

\bibitem{juho}
J.S. Lintuvuori, D. Marenduzzo, K. Stratford, and M.E. Cates,
\textit{J. Mater. Chem.} \textbf{20}, 10547 (2010).

@@

\bibitem{oliver-bp3}
O. Henrich, K. Stratford, M.~E. Cates, D. Marenduzzo,
{\it Phys. Rev. Lett.} {\bf 106}, 107802 (2011).

@@

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=0.95\columnwidth]{support-fig7.pdf}
\end{center}
\caption{Panels A and B as for Fig. 3, but with the disclination lines
removed for clarity. Panel C shows the situation when the field is
switched off: a residual anisotropic structure remains which reflects the
structure induced by the electric field. The electric field direction
when switched on (panel B) is into the plane of the paper.}
\end{figure}

@@

{\bf Supplementary Movie 1:} 
The final state of the simulation as shown in Fig.~1A. The movie
rotates the configuration to reveal more clearly its structure.\\

{\bf Supplementary Movie 2:} 
The final state of the simulation as shown in Fig.~1B. The movie
rotates the configuration to reveal more clearly its structure.\\

{\bf Supplementary Movie 3:} 
The final state of the simulation as shown in Fig.~1C. The movie
rotates the configuration to reveal more clearly its structure.\\

{\bf Supplementary Movie 4:} 
The final state of the simulation as shown in Fig.~1D. The movie
rotates the configuration to reveal more clearly its structure.\\

{\bf Supplementary Movie 5:} 
The configuration after the initial relaxation phase in Fig.~3. The movie
shows a rotation of the configuration. The distribution of the colloidal
particles is isotropic, with the particles forming clusters and having a
tendency to occupy positions at the intersection of the BPIII disclination
lines.\\

{\bf Supplementary Movie 6:} 
Configuration in a homogeneous electric field with strength
corresponding to ${\cal E}=0.8$. The movie
shows a rotation of the configuration.
A hexagonal pattern emerges where the particles form chains
situated at the vertices of the honeycomb-like structure. \\
